---
title: "BOOTEI on the Soybean data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BOOTEI on the Soybean data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```



## Overview

This vignette illustrates how **BOOTEI** refines classical permutation tests when the
primary test statistic is discrete and ties are likely.

We demonstrate:

1. A Pearson chi-square test of independence within a single disease class
   of the Soybean dataset.
2. A two-sample Mann--Whitney comparison across two disease classes on an
   ordinally encoded attribute.

We compare the baseline add-one Monte Carlo permutation p-value
(obtained by setting `B = 1`) with the BOOTEI-refined p-value
(using an ensemble plan with `B = 200`).

## Data

We use the `Soybean` dataset from the `mlbench` package.

```{r data}
library(bootei)
library(mlbench)

data("Soybean", package = "mlbench")

dd <- Soybean
dd$Class <- droplevels(dd$Class)   # stable indexing of class labels
classes  <- levels(dd$Class)
```


## Helpers

The following helpers mirror the preprocessing used in the full automated
scan scripts.


```{r helpers}
to_numeric_ordinal <- function(z) {
  if (is.factor(z) || is.character(z)) {
    suppressWarnings(as.numeric(as.character(z)))
  } else {
    suppressWarnings(as.numeric(z))
  }
}

prep_cat <- function(z) {
  if (is.factor(z)) droplevels(z) else droplevels(factor(z))
}
```


## Example 1: Pearson chi-square (within-class independence)
text
Copy code


```{r chi2-example}
sub <- dd[dd$Class == classes[3], , drop = FALSE]

x0 <- sub[[3]]
y0 <- sub[[31]]

ok <- !(is.na(x0) | is.na(y0))
x  <- prep_cat(x0[ok])
y  <- prep_cat(y0[ok])

table(x, y)
```


```{r chi2-pvalues}
p_perm_chi2 <- bootei(
  x, y,
  B = 1L, R = 5000L, perm_seed = 910L,
  test = "chisq"
)$p.value

p_bootei_chi2 <- bootei(
  x, y,
  B = 200L, R = 5000L, perm_seed = 910L,
  test = "chisq"
)$p.value

p_perm_chi2
p_bootei_chi2
```


## Example 2: Mann--Whitney (two-sample comparison across classes)
text
Copy code



```{r mw-example}
x <- to_numeric_ordinal(dd[dd$Class == classes[15], 10])
y <- to_numeric_ordinal(dd[dd$Class == classes[19], 10])

x <- x[!is.na(x)]
y <- y[!is.na(y)]

c(nx = length(x), ny = length(y))
c(uniq_x = length(unique(x)), uniq_y = length(unique(y)))
```


```{r mw-pvalues}
p_perm_mw <- bootei(
  x, y,
  B = 1L, R = 5000L, perm_seed = 910L,
  test = "mannwhitney"
)$p.value

p_bootei_mw <- bootei(
  x, y,
  B = 200L, R = 5000L, perm_seed = 910L,
  test = "mannwhitney"
)$p.value

p_perm_mw
p_bootei_mw
```



## Notes on reproducibility

Fixing the permutation seed ensures reproducibility of the Monte Carlo
permutation distribution. BOOTEI differs from the baseline permutation test
only when the primary statistic produces ties; when ties are rare, the two
procedures coincide almost exactly.




